---
title: "databusters_"
author: "YH"
date: "2025-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(tseries)
```

```{r}
library(readr)
library(readxl)
library(stringr)
library(lubridate)
library(tidyverse)
library(dplyr)
library(dynlm)
library(vars)
library(tseries)
library(forecast)
library(ggcorrplot)
library(rlang)
library(reshape2)
library(fanplot)
```


# Exploring the data
```{r explore, echo=FALSE}
vars = read_csv("Variable Description.csv", show_col_types = FALSE)
df = read_csv("Quarterly Data.csv", show_col_types = FALSE)
```



Selecting indicators that we want to use only
```{r}
df_select <- df[-(1:2),] 
df_select <- df_select %>% 
  dplyr:: select(sasdate, GDPC1, UNRATE, PCECC96,CPILFESL) %>%
  mutate(sasdate = dmy(sasdate))

```


### Load and prepare data
```{r}
start_year <- year(min(df_select$sasdate, na.rm = TRUE)) 
start_quarter <- quarter(min(df_select$sasdate, na.rm = TRUE))
ts_data <- ts(df_select[, c("GDPC1", "UNRATE", "PCECC96", "CPILFESL")], start = c(start_year,start_quarter), frequency = 4)
```


### Data Exploration 
```{r}
ggplot(df_select, aes(x = sasdate, y = GDPC1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) + 
  labs(title = "General GDP Trend", x = "Year", y = "GDPC1") +
  theme_minimal() + 
  scale_x_date(limits = c(as.Date("1980-01-03"), as.Date("2024-01-09"))) 

```


```{r}
# Correlation matrix calculation excluding the date column
cor_matrix <- cor(df_select[, -1])  # Exclude the first column (sasdate)

# Reshape the correlation matrix for ggplot2
library(reshape2)
cor_long <- melt(cor_matrix)

# Set the order of the variables for both x and y axes to match
cor_long$Var1 <- factor(cor_long$Var1, levels = colnames(cor_matrix))
cor_long$Var2 <- factor(cor_long$Var2, levels = colnames(cor_matrix))

# Plot the correlation heatmap with consistent variable order
library(ggplot2)
ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile() +  # Create tiles
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +  # Color gradient
  theme_minimal() +  # Clean theme
  labs(title = "Correlation Matrix Heatmap", x = "Variables", y = "Variables") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```

### Splitting Train/Test Sets 
```{r}
train_data <- window(ts_data, start=1959, end=2012) # Use all but last 4 points for training
test_data <-  window(ts_data, start=2012) # Last 4 points as test set
test_data <- test_data[-1,]
```


### Determining optimal lag lengths
Determining p
```{r}
p_values <- 1:10

# Create an empty list to store AIC values
aic_values <- numeric(length(p_values))

# Loop over different values of p
for (p in p_values) {
  # Construct the formula dynamically
  formula_str <- paste0("GDPC1 ~ ", 
                        paste0("L(GDPC1, ", 1:p, ")", collapse = " + "))  
  model <- dynlm(as.formula(formula_str), data = train_data)
  aic_values[p] <- AIC(model)
}

# Create a data frame to display AIC values
aic_results <- data.frame(p = p_values, AIC = aic_values)

# Print AIC values for comparison
print(aic_results)

#Plot a graph to see which lag will give the lowest AIC
plot(aic_results$p, aic_results$AIC, type="b", pch=19, col="blue",
     xlab="Number of Lags for GDPC1", ylab="AIC", main="AIC for Different GDPC1 Lags")

```


Choosing UNRATE
number of lags: 2
```{r}
# Define the range of lags for UNRATE
unrate_lags <- 1:10  

# Create an empty list to store AIC values
aic_values2 <- numeric(length(unrate_lags))

# Loop over different lag values for UNRATE
for (lag in unrate_lags) {
  # Construct the formula dynamically
  formula_str <- paste0("GDPC1 ~ ", 
                        paste0("L(GDPC1, ", 1:10, ")", collapse = " + "),  # Fixed lags for GDPC1
                        " + ", 
                        paste0("L(UNRATE, ", 1:lag, ")", collapse = " + "))  # Varying lags for UNRATE
  model <- dynlm(as.formula(formula_str), data = train_data)
  aic_values2[lag] <- AIC(model)
}

# Create a data frame to display AIC values
aic_results2 <- data.frame(UNRATE_Lags = unrate_lags, AIC = aic_values2)

# Print the AIC values for comparison
print(aic_results2) 

#Plot a graph to see which lag will give the lowest AIC
plot(aic_results2$UNRATE_Lags, aic_results2$AIC, type="b", pch=19, col="blue",
     xlab="Number of Lags for UNRATE", ylab="AIC", main="AIC for Different UNRATE Lags")

```

Choosing PCECC96
number of lags: 3
```{r}
# Define the range of lags for PCECC96
pcecc96_lags <- 1:10  

# Create an empty list to store AIC values
aic_values3 <- numeric(length(pcecc96_lags))

# Loop over different lag values for PCECC96
for (lag in pcecc96_lags) {
  # Construct the formula dynamically
  formula_str <- paste0("GDPC1 ~ ", 
                        paste0("L(GDPC1, ", 1:10, ")", collapse = " + "),  # Fixed lags for GDPC1
                        " + ", 
                        paste0("L(UNRATE, ", 1:3, ")", collapse = " + "),  # Fixed lags for UNRATE
                        " + ", 
                        paste0("L(PCECC96, ", 1:lag, ")", collapse = " + "))  # Varying lags for PCECC96
  model <- dynlm(as.formula(formula_str), data = train_data)
  aic_values3[lag] <- AIC(model)
}

# Create a data frame to display AIC values
aic_results3 <- data.frame(PCECC96_Lags = pcecc96_lags, AIC = aic_values3)

# Print the AIC values for comparison
print(aic_results3)

#Plot a graph to see which lag will give the lowest AIC
plot(aic_results3$PCECC96_Lags, aic_results3$AIC, type="b", pch=19, col="blue",
     xlab="Number of Lags for PCECC96", ylab="AIC", main="AIC for Different PCECC96 Lags")

```

Choosing CPILFESL
number of lags: 1
```{r}
# Define the range of lags for CPILFESL
CPILFESL_lags <- 1:10  

# Create an empty list to store AIC values
aic_values4 <- numeric(length(CPILFESL_lags))

# Loop over different lag values for PCECC96
for (lag in CPILFESL_lags) {
  # Construct the formula dynamically
  formula_str <- paste0("GDPC1 ~ ", 
                        paste0("L(GDPC1, ", 1:10, ")", collapse = " + "),  # Fixed lags for GDPC1
                        " + ", 
                        paste0("L(UNRATE, ", 1:3, ")", collapse = " + "),  # Fixed lags for UNRATE
                        " + ", 
                        paste0("L(PCECC96, ", 1:4, ")", collapse = " + "), # Fixed lags for PCECC96 
                        " + ", 
                        paste0("L(CPILFESL, ", 1:lag, ")", collapse = " + "))  #Varying lags for CPILFESL
  model <- dynlm(as.formula(formula_str), data = train_data)
  aic_values4[lag] <- AIC(model)
}

# Create a data frame to display AIC values
aic_results4 <- data.frame(CPILFESL_Lags = CPILFESL_lags, AIC = aic_values4)

# Print the AIC values for comparison
print(aic_results4)

#Plot a graph to see which lag will give the lowest AIC
plot(CPILFESL_lags, aic_results4$AIC, type="b", pch=19, col="blue",
     xlab="Number of Lags for CPILFESL_lags", ylab="AIC", main="AIC for Different CPILFESL Lags")

```



### Final model
```{r}
final_model <- dynlm(GDPC1 ~ L(GDPC1, 1) + L(GDPC1, 2) + L(GDPC1, 3) + L(GDPC1, 4) + L(GDPC1, 5) + L(GDPC1, 6) + L(GDPC1, 7) + L(GDPC1, 8) +L(GDPC1, 9) +L(GDPC1, 10) + L(UNRATE, 1) + L(UNRATE, 2) +  L(UNRATE, 3) + L(PCECC96, 1)+ L(PCECC96, 2)+ L(PCECC96, 3) + L(PCECC96, 3) + L(CPILFESL,1), data = train_data)
```


Using arima to predict values for UNRATE
```{r}
data <- ts_data[,2]
model <- auto.arima(data)  # Simpler AR(1) model
summary(model)

value_UNRATE <- forecast(model, h=5)
print(value_UNRATE)
```


Using arima to predict values for PCECC96
```{r}
data2 <- ts_data[,3]
model2 <- auto.arima(data2)  # Simpler AR(1) model
summary(model2)

value_PEC <- forecast(model2, h=5)
print(value_PEC)

```

Using arima to predict values for CPILFESL
```{r}
data3 <- ts_data[,4]
model3 <- auto.arima(data3)  # Simpler AR(1) model
summary(model3)

value_CPILFESL <- forecast(model2, h=5)
print(value_CPILFESL)

```



### Extracting the point forecasts values
```{r}
point_forecasts_pec <- value_PEC$mean
point_forecasts_values <- as.numeric(point_forecasts_pec)

point_forecasts_UNRATE <- value_UNRATE$mean
point_forecasts_values_ <- as.numeric(point_forecasts_UNRATE) 

point_forecasts_CPILFESL <- value_CPILFESL$mean
point_forecasts_values_1 <- as.numeric(point_forecasts_CPILFESL) 
# Create a new data frame and add the new point forecast values to ts_data
new_rows <- cbind(c(0,0,0,0,0),point_forecasts_values_, point_forecasts_values,point_forecasts_values_1)

new_data_frame <- rbind(ts_data, new_rows)
```


# Forecasting
```{r}
#Forecast for 5 periods (because our data is missing 2024 Q4 for some reason)
forecast_values <- predict(final_model, n.ahead = 5, newdata=data.frame(new_data_frame))
print(forecast_values)
#str(forecast_values)

forecast_values_gdpc1 <- as.numeric(forecast_values[264:268])
future_dates <- seq(as.Date("2024-12-01"), as.Date("2025-12-01"), by = "quarter")

quarter_year_labels <- paste0(format(future_dates, "%Y"), quarters(future_dates))

# Combine the forecasted values, dates, and quarter-year labels into a data frame
plot_data <- data.frame(Date = future_dates, Quarter_Year = quarter_year_labels, GDP = forecast_values_gdpc1)


old_gdp_data <- df_select %>%
  dplyr::select(sasdate, GDPC1) %>%
  slice(257:n()) %>%
  #mutate(sasdate = as.Date(sasdate, format = "%Y-%d-%m")) %>%
  rename(Date = sasdate) %>%
  mutate(Quarter_Year = paste0(format(Date, "%Y"), "Q", 
                                      ceiling(as.numeric(format(Date, "%d")) / 3))) %>%
  rename(GDP = GDPC1)%>%
  dplyr::select(1, 3, 2, everything())

plot_data = rbind(old_gdp_data,plot_data)


```


```{r}
ggplot(plot_data, aes(x = Quarter_Year, y = GDP, group = 1)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) + 
  labs(title = "Forecasted GDP Trend", x = "Quarter-Year", y = "GDP") +
  theme_minimal()
  #scale_x_discrete(limits = quarter_year_labels)

```


# Evaluating Our Model 
```{r}
forecast_values <- predict(final_model, newdata = data.frame(test_data))
accuracy(forecast_values, test_data[,1])
```



